Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
revert and try again

## Create Patient-Level Summary Variables

```{r create-patient-summaries}
# Create patient-level summary variables that capture severity without collinearity
# These aggregates across eyes provide information about patient-level disease burden

df_with_patient_summaries <- df_for_imputation |>
  group_by(e_mrn_deidentified) |>
  mutate(
    n_eyes = n(),

    # Patient-level baseline summaries
    patient_worst_baseline = max(baseline_VA_logMAR, na.rm = TRUE),
    patient_best_baseline = min(baseline_VA_logMAR, na.rm = TRUE),
    patient_mean_baseline = mean(baseline_VA_logMAR, na.rm = TRUE),

    # Patient-level intermediate summaries  
    patient_worst_intermediate = max(intermediate_VA_logMAR, na.rm = TRUE),
    patient_best_intermediate = min(intermediate_VA_logMAR, na.rm = TRUE),
    patient_mean_intermediate = mean(intermediate_VA_logMAR, na.rm = TRUE),

    # Asymmetry measures (for bilateral patients)
    baseline_asymmetry = if_else(n_eyes == 2,
                                 abs(baseline_VA_logMAR[1] - baseline_VA_logMAR[2]),
                                 NA_real_),
    intermediate_asymmetry = if_else(n_eyes == 2,
                                    abs(intermediate_VA_logMAR[1] - intermediate_VA_logMAR[2]),
                                    NA_real_)
  ) |>
  ungroup() |>
  mutate(
    # Handle Inf values from all-NA groups
    across(starts_with("patient_worst"), ~if_else(is.infinite(.), NA_real_, .)),
    across(starts_with("patient_best"), ~if_else(is.infinite(.), NA_real_, .)),
    across(starts_with("patient_mean"), ~if_else(is.nan(.), NA_real_, .))
  )

# Log summary statistics
patient_summary_stats <- df_with_patient_summaries |>
  summarise(
    total_patients = n_distinct(e_mrn_deidentified),
    total_eyes = n(),
    bilateral_patients = sum(n_eyes == 2) / 2,
    single_eye_patients = sum(n_eyes == 1),

    # Check data availability
    has_patient_worst_baseline = sum(!is.na(patient_worst_baseline)),
    has_patient_best_baseline = sum(!is.na(patient_best_baseline)),
    has_baseline_asymmetry = sum(!is.na(baseline_asymmetry)),

    # Average asymmetry for bilateral patients
    mean_baseline_asymmetry = mean(baseline_asymmetry, na.rm = TRUE),
    mean_intermediate_asymmetry = mean(intermediate_asymmetry, na.rm = TRUE)
  )

log_message(paste0("Patient-level summaries created:"))
log_message(paste0("  Total patients: ", patient_summary_stats$total_patients))
log_message(paste0("  Bilateral patients: ", patient_summary_stats$bilateral_patients))
log_message(paste0("  Mean baseline asymmetry (bilateral): ", 
                  round(patient_summary_stats$mean_baseline_asymmetry, 3)))

# Check correlations between patient summaries and individual eyes
correlation_check <- df_with_patient_summaries |>
  summarise(
    # These should be high but not perfect (avoiding collinearity)
    cor_baseline_worst = cor(baseline_VA_logMAR, patient_worst_baseline, use = "complete.obs"),
    cor_baseline_best = cor(baseline_VA_logMAR, patient_best_baseline, use = "complete.obs"),
    cor_baseline_mean = cor(baseline_VA_logMAR, patient_mean_baseline, use = "complete.obs")
  )

log_message(paste0("Correlations with patient summaries - Worst: ", 
                  round(correlation_check$cor_baseline_worst, 3),
                  ", Best: ", round(correlation_check$cor_baseline_best, 3),
                  ", Mean: ", round(correlation_check$cor_baseline_mean, 3)))

# Use this enhanced dataset for imputation
df_for_imputation <- df_with_patient_summaries
```

### Patient-Level Summary Statistics

```{r patient-summary-table}
# Display patient distribution
patient_dist <- data.frame(
  Category = c("Total patients", "Total eyes", "Bilateral patients", "Single eye patients"),
  N = c(patient_summary_stats$total_patients, 
        patient_summary_stats$total_eyes,
        patient_summary_stats$bilateral_patients,
        patient_summary_stats$single_eye_patients)
)
kable(patient_dist, caption = "Patient/Eye Distribution")

# Display summary variable availability
summary_availability <- data.frame(
  Variable = c("Patient worst baseline", "Patient best baseline", "Baseline asymmetry"),
  N_Available = c(patient_summary_stats$has_patient_worst_baseline,
                 patient_summary_stats$has_patient_best_baseline,
                 patient_summary_stats$has_baseline_asymmetry),
  Percent = round(100 * c(patient_summary_stats$has_patient_worst_baseline,
                          patient_summary_stats$has_patient_best_baseline,
                          patient_summary_stats$has_baseline_asymmetry) / 
                 patient_summary_stats$total_eyes, 1)
)
kable(summary_availability, caption = "Patient Summary Variable Availability")

# Display asymmetry statistics for bilateral patients
asymmetry_stats <- data.frame(
  Measure = c("Mean baseline asymmetry", "Mean intermediate asymmetry"),
  Value = round(c(patient_summary_stats$mean_baseline_asymmetry,
                  patient_summary_stats$mean_intermediate_asymmetry), 3)
)
kable(asymmetry_stats, caption = "Visual Acuity Asymmetry (Bilateral Patients)")
```

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr",  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr

  # Patient-level summary variables (capture severity without collinearity)
  "n_eyes",  # Number of eyes per patient
  "patient_worst_baseline",  # Max baseline VA across eyes
  "patient_best_baseline",   # Min baseline VA across eyes
  "patient_worst_intermediate",  # Max intermediate VA across eyes
  "patient_best_intermediate"   # Min intermediate VA across eyes
  # Note: asymmetry variables excluded as they're only for bilateral patients
)

@@

# Patient-level summary variables - don't impute (they're derived)
custom_methods["n_eyes"] <- ""
custom_methods["patient_worst_baseline"] <- ""
custom_methods["patient_best_baseline"] <- ""
custom_methods["patient_worst_intermediate"] <- ""
custom_methods["patient_best_intermediate"] <- ""

@@

# PATIENT-LEVEL SUMMARY PREDICTORS
# These capture patient-level severity without causing collinearity

# Patient summaries should not be predicted (they're derived)
if ("patient_worst_baseline" %in% rownames(pred_matrix)) {
  pred_matrix["patient_worst_baseline", ] <- 0
  pred_matrix["patient_best_baseline", ] <- 0
  pred_matrix["patient_worst_intermediate", ] <- 0
  pred_matrix["patient_best_intermediate", ] <- 0
  pred_matrix["n_eyes", ] <- 0
}

# Patient baseline summaries can predict intermediate and outcome
# But NOT baseline itself (would be circular)
if ("patient_worst_baseline" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "patient_worst_baseline"] <- 0  # Don't use to predict baseline
  pred_matrix["intermediate_VA_logMAR", "patient_worst_baseline"] <- 1
  pred_matrix["outcome_VA_logMAR", "patient_worst_baseline"] <- 1

  pred_matrix["baseline_VA_logMAR", "patient_best_baseline"] <- 0  # Don't use to predict baseline
  pred_matrix["intermediate_VA_logMAR", "patient_best_baseline"] <- 1
  pred_matrix["outcome_VA_logMAR", "patient_best_baseline"] <- 1
}

# Patient intermediate summaries can predict outcome only
# NOT baseline or intermediate (temporal constraint)
if ("patient_worst_intermediate" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "patient_worst_intermediate"] <- 0
  pred_matrix["intermediate_VA_logMAR", "patient_worst_intermediate"] <- 0  # Don't predict intermediate
  pred_matrix["outcome_VA_logMAR", "patient_worst_intermediate"] <- 1

  pred_matrix["baseline_VA_logMAR", "patient_best_intermediate"] <- 0
  pred_matrix["intermediate_VA_logMAR", "patient_best_intermediate"] <- 0
  pred_matrix["outcome_VA_logMAR", "patient_best_intermediate"] <- 1
}

# n_eyes can be informative for missingness patterns but shouldn't predict values directly
if ("n_eyes" %in% colnames(pred_matrix)) {
  pred_matrix[, "n_eyes"] <- 0  # Don't use as predictor to avoid collinearity
}

log_message("Patient-level summary predictors configured")

@@

@@

# Variables that will ACTUALLY be imputed (have missing data AND a method)
# Exclude patient summary variables even if they have missing data
vars_to_impute <- final_methods |>
  filter(Will_Be_Imputed) |>
  filter(!Variable %in% c("n_eyes", "patient_worst_baseline", "patient_best_baseline",
                          "patient_worst_intermediate", "patient_best_intermediate")) |>
  pull(Variable)
```

>>>>

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr
)

@@

@@

@@

# Check for perfect collinearity between intermediate and outcome for submitted cohort
collinearity_check <- df_mice |>
  filter(cohort_id == "submitted") |>
  summarise(
    n = n(),
    n_with_both = sum(!is.na(intermediate_VA_logMAR) & !is.na(outcome_VA_logMAR)),
    cor_intermediate_outcome = cor(intermediate_VA_logMAR, outcome_VA_logMAR, use = "complete.obs"),
    are_identical = all(intermediate_VA_logMAR == outcome_VA_logMAR, na.rm = TRUE)
  )

log_message(paste0("Submitted cohort - Correlation between intermediate and outcome: ",
                  round(collinearity_check$cor_intermediate_outcome, 3)))
log_message(paste0("Submitted cohort - Are intermediate and outcome identical? ",
                  collinearity_check$are_identical))

# Log which predictors are set for outcome_VA_logMAR
outcome_predictors <- names(which(pred_matrix["outcome_VA_logMAR", ] == 1))
log_message(paste0("Predictors for outcome_VA_logMAR (", length(outcome_predictors), "): ",
                  paste(outcome_predictors, collapse = ", ")))

@@

# Variables that will ACTUALLY be imputed (have missing data AND a method)
vars_to_impute <- final_methods |>
  filter(Will_Be_Imputed) |>
  pull(Variable)
```
